Coherence between subjective feeling and facial expression: A comparison of human and automatic recognition

### 2019-04-05 ACII_2019_paper/ACII_2019_paper.tex
latex

\hypersetup{
            pdftitle={Are facial expressions the genuine display of individuals' subjective feeling? A comparison with human and automatic recognition.},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Are facial expressions the genuine display of individuals' subjective
feeling? A comparison with human and automatic recognition.}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
While it has been taken for granted in the development of several
automatic facial expression recognition tools, the question of the link
between subjective feelings and facial expressions is still a subject of
debate. On one hand the behaviorist approach conceives emotions as
genetically hardwired and therefore being genuinely displayed through
facial expressions. On the other hand the constructivist approach
conceives emotions are socially constructed and facial expression as
social messages that are not related to emotions. In order to evaluate
the link between the subjective feeling of emotions and their
recognition based on facial expression, 232 videos of participants
recruited to perform an emotion elicitation task were annotated by 1383
human observers as well as by an automatic facial expression classifier.
The results show a low accuracy of human observers and of the automatic
classifier to infer from the facial expression the subjective feeling of
the participants recorded. Based on these results, the hypothesis of
genetically hardwired emotion genuinely display is difficult to support
whereas the idea of emotion socially constructed and facial expression
as display of social messages appears to be more likely. Then, the way
to infer emotional and mental state based on facial expressions should
be questioned.
\end{abstract}

@@

With the development of commercial automatic facial expression
recognition tools (see Dupré et al. 2018 for a non-exhaustive list of
available tools), industries and governments are gradually implementing
this technology in order to track humans' emotions in various scenarios
(e.g., marketing, healthcare, automotive to name a few). Beside the
ethical question of measuring human emotions, these tools are
challenging psychological theories about inferring emotions from facial
expressions. When automatic facial expression recognition tools are used
to measure human emotions, they rest on the premise that facial
expressions provide a direct access to expressers' emotions. However the
link between emotion felt and facial expressions is far from being
established and still remain a hot topic in psychology research.

\hypertarget{the-link-between-emotion-and-facial-expression-thought-the-behaviorist-approach}{%
\subsection{The link between emotion and facial expression thought the
behaviorist
approach}\label{the-link-between-emotion-and-facial-expression-thought-the-behaviorist-approach}}

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} (Darwin 1872), facial
expression are conceived as a genuine display of individuals inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory (BET) which states that a set of six or seven emotions
are universally displayed and are genetically hardwired not only in
humans (Ekman 1992) but also in different animal species (Waal 2019).
According to this view, ``when emotions are aroused by perception of a
social event, a set of central commands produce patterned
emotion-specific changes in multiple systems, including {[}\ldots{}{]}
facial expressions.'' (Ekman 2007, p49). Even if this theory obtained a
popular support, it fails to explain how individuals can feel emotions
without expressing them and how individuals can express emotions without
feeling them.

\hypertarget{the-link-between-emotion-and-facial-expression-thought-the-social-constructivist-approach}{%
\subsection{The link between emotion and facial expression thought the
social constructivist
approach}\label{the-link-between-emotion-and-facial-expression-thought-the-social-constructivist-approach}}

Detractors of the Basic Emotion Theory are perceiving emotion not as
genetically hardwired but as a learnt association between a given
situation and an appropriate response (Averill 1980,
@barrett2017emotions). For the tenants of the constructivist approach,
emotions are ``concepts'' based on past experiences and which are ``a
collection of embodied, whole brain representations that predict what is
about to happen in the sensory environment, what the best action is to
deal with impending events, and their consequences for allostasis''
(Barrett 2017b, p12). Following this assumption, faces are used as tools
to display signals in social interactions (Crivelli and Fridlund 2018).
These signals can convey individuals' motivations and readiness (Frijda
and Tcherkassof 1997) or social messages (Fridlund and Rosenberg 1995).

This current paper investigates the link between the subjective feeling
of emotions and their recognition from facial expressions. If emotions
are hardwired, individuals emotional subjective feeling should be
correlated to the recognition of facial expressions from both human
observers and automatic classifiers whereas if emotions are social
constructs, no correlation between subjective feeling and facial
expression recognition should be observed.

@@

To evaluate the link between subjective feeling of emotions and their
recognition from facial expressions, participants were recruited to
perform an emotion elicitation task while their facial expression was
video recorded. Then, the videos was shown to human observer and
analysed by an automatic classifier in order to identify which emotion
was displayed.

@@

For the emotion elicitation experiment, 358 French participants (182
females, 176 males, \emph{M}age = 47.9, \emph{SD}age = 9.2) were
recruited to perform one out of 11 emotions elicitation tasks designed
to trigger a positive, negative or neutral emotional state (see
Tcherkassof et al. 2013 for a description of tasks and procedure).
Participants' face were recorded using an hidden camera resulting 358
front facing 768x576 videos varying from 1s to 1479s.

After their emotion elicitation task the participants had to rate their
emotional state during the task on a likert scale from 0 (``not at
all'') to 5 (``strongly'') the six ``basic'' emotions (i.e.,
\emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) as well as six ``non-basic''
emotions (i.e., \emph{pride}, \emph{curiosity}, \emph{boredom},
\emph{shame}, \emph{humiliation}, and \emph{disappointment}).

Finally, a debriefing session was perform to ensure that participants
were not durably affected by the emotion elicitation task. All the
participants gave their agreement on their data and video to be
processed for research purpose only.

@@

For the human facial expression recognition method, 1383 student
participants were recruited to annotate 232 out of the 358 video,
therefore only the 232 annotated videos will be analysed in this paper.
Each participants had to annotate between 1 and 177 videos resulting
that each video was annotated 29 times on average (\emph{SD} = 12).

The annotation of facial expressions was performed on-site using
\emph{Oudjat}, a software for designing video annotation experiments
(Dupré et al. 2015). For each video, the annotation procedure hat two
steps. First, the participants had to identify the emotional sequences
by pressing the space bar of their keyboard to indicate the beginning
and the end of the emotional sequences while watching the video. Second,
the participants watched each emotional sequence previously identified
and had to label the sequence using one of the 12 emotions proposed
including six ``basic'' emotions (i.e., \emph{anger}, \emph{disgust},
\emph{fear}, \emph{happiness}, \emph{surprise} and \emph{sadness}) and
six ``non-basic'' emotions (i.e., \emph{pride}, \emph{curiosity},
\emph{boredom}, \emph{shame}, \emph{humiliation}, and
\emph{disappointment}). They also had the possibility to indicate that
the sequence was not expressing one of the proposed emotion.

This annotation procedure results in a uni-dimensional time-series for
each video per human observer identifying for each second of the video
which emotion was recognized. Then time-series corresponding to the same
video were aggregated to calculate the proportion of human observers for
each second of the video per emotional label. The sum of each label
proportion per second was used as a score to determine which labels
corresponds to the overall video (i.e., the highest score). In case of
more than one label having the maximum value, the emotion is described
as undetermined.

@@

The 232 annotated video were processed with Affdex (SDK v3.4.1). Affdex
is an automatic facial expression recognition classifier developed and
distributed by Affectiva is a spin-off company resulting from the
research activities of MIT media lab created in 2009 (McDuff et al.
2016). Affdex's algorithm uses Histogram of Oriented Gradient (HOG)
features and Support Vector Machine (SVM) classifiers in order to
recognize facial expressions. For each video frame, Affdex identify the
probability of the face as expressing one of the six ``basic'' emotions
(i.e., \emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) as well as additional psychological
states such as \emph{valence}, \emph{engagement} or \emph{contempt}, and
facial features such as \emph{cheek raise}, \emph{eye widen} or
\emph{jaw drop}.

To determine which of the six ``basic'' emotion can be used to identify
each video, the recognition probability for each label by frame was
converted into odd ratio by frame (Dente et al. 2017). The highest sum
of each odd ratio time-series defines the label recognized by the
automatic classifier.

@@

Whereas the self-reports, the human annotations and the automatic
recognition include data on ``non-basic'' emotions and features, the
analysis is performed using only the six ``basic'' emotions in order to
compare them. The maximum score for self-reports, human annotations and
automatic recognition is used to label the video. In case of more than
one label obtaining the maximum value, the video is labeled as
undetermined.

@@

Emotions self-reported as being characteristic of the elicitation are
compared with the emotion recognized by the human observers in a
confusion matrix (Figure \ref{fig:confusionMatrix_sr_hr}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_hr-1.pdf}
\caption{Confusion matrix of between the emotion self-reported as being
characteristic of the elicitation with the emotion recognized by the
human observers.}
\end{figure}

The result of the confusion matrix show a low agreement between emotion
felt during the elicitation and emotion recognized by the human
annotators (Accuracy = 0.27, 95\%CI{[}0.21,0.33{]}; Kappa = 0.11) except
for \emph{happiness} (15.2\%), \emph{surprise} (8.26\%) and
\emph{disgust} (1.74\%). Sensitivity, specificity, precision and F1
score for each emotion can be found Table
\ref{tab:confusionTable_sr_hr}. Interestingly human annotators seem to
recognize as \emph{surprise} videos in which \emph{happiness} was the
highest self-reported emotion (10.9\%), and in a lower instance
\emph{happiness} videos in which \emph{surprise} was the highest
self-reported emotion (4.78\%).

@@

\caption{\label{tab:confusionTable_sr_hr}Agreement accuracy metrics for each emotion. }
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
Emotion & Sensitivity & Specificity & Precision & F1\\
\midrule
anger & 0.43 & 0.92 & 0.14 & 0.21\\
disgust & 1.00 & 0.93 & 0.20 & 0.33\\
fear & 0.00 & 0.96 & 0.00 & \textit{na.}\\
happiness & 0.49 & 0.73 & 0.45 & 0.47\\
sadness & 0.08 & 0.91 & 0.05 & 0.06\\
surprise & 0.44 & 0.67 & 0.23 & 0.31\\
undetermined & 0.00 & 0.99 & 0.00 & \textit{na.}\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} Note. \textit{na.} values are produced when not enough}\\
\multicolumn{5}{l}{data are available to compute accuracy indicators.}\\
\end{tabu}
\end{table}

However, the self-report show a very high proportion of undetermined
emotional states which reveals not only the possibility of the emotion
elicitation tasks to trigger more than one emotion but also the
potential limit of using 6-points likert scales for which the
participants can easily score to the maximum for more than one emotion.

@@

As in the previous analysis, emotions self-reported as being
characteristic of the elicitation are compared with the emotion
recognized by the automatic classifier in a confusion matrix (Figure
\ref{fig:confusionMatrix_sr_ar}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_ar-1.pdf}
\caption{Confusion matrix of between the emotion self-reported as being
characteristic of the elicitation with the emotion recognized by the
automatic classifier.}
\end{figure}

Results obtained for the comparison between emotions self-reported and
recognized by the automatic classifier are similar to the ones with
human observers (Table \ref{tab:confusionTable_sr_ar}). Overall there is
a low agreement between emotion self-reported and emotion recognized by
the automatic classifier (Accuracy = 0.19, 95\%CI{[}0.14,0.24{]}; Kappa
= 0.05) except for \emph{happiness} (13.9\%) and \emph{surprise}
(3.48\%). Surprisingly the automatic classifier incorrectly recognized
as \emph{disgust} an important proportion of videos in which
\emph{happiness} was the highest self-reported emotion (7.83\%). In
parallel, the automatic classifier recognized as \emph{happiness} and
\emph{disgust} videos in which \emph{surprise} was the highest
self-reported emotion (respectively 5.22\% and 6.09\%).

@@

\caption{\label{tab:confusionTable_sr_ar}Agreement accuracy metrics for each emotion.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
Emotion & Sensitivity & Specificity & Precision & F1\\
\midrule
anger & 0.00 & 0.89 & 0 & \textit{na.}\\
disgust & 0.25 & 0.70 & 0.01 & 0.03\\
fear & 0.00 & 0.93 & 0 & \textit{na.}\\
happiness & 0.44 & 0.71 & 0.41 & 0.43\\
sadness & 0.15 & 0.94 & 0.14 & 0.15\\
surprise & 0.19 & 0.89 & 0.28 & 0.22\\
undetermined & 0.00 & 1.00 & \textit{na.} & \textit{na.}\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} Note. \textit{na.} values are produced when not enough data are}\\
\multicolumn{5}{l}{available to compute accuracy indicators.}\\
\end{tabu}
\end{table}

A comparable explanation can be provided as the level of undetermined
emotion are very high for the self reports.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Despite being one on the most investigated question in affective
science, the link between emotion felt and facial expression is a hot
topic and no clear evidence have been found to definitely answer it.
However, with the growing interest of industries and government to
monitor individual's psychological states, evidences are showing that
facial expressions are in reality not expressing emotions (McKeown
2013). This research aimed to provide some empirical data to the
question. The subjective feeling of participants was compared with human
recognition on one side and automatic recognition on the other side. The
results reveals a low accuracy for both humans and automatic classifier
to accurately identify the inner emotional states of these individuals
based on their facial expressions.

Some limitations to this process should be stated over the use of
self-reports to evaluate individual's subjective feelings. Accessing to
the inner subjective feeling can be biased if not impossible. Moreover
the laboratory setting can trigger ambiguous and ``non-basic'' emotions
which were not analysed in this research. The procedure use for human
annotation can also be incriminated. Instead of asking the human
annotators to provide an unique label, a more subtle approach was chosen
to mimic results provided by the automatic classifier. In this regard,
the results of the human annotation could have been more ambiguous
because it is not the natural way that people are inferring human
emotions. Finally, the automatic classifier algorithm can also be
problematic. Based on training datasets which are most of the time using
prototypical facial expression of the ``basic'' emotions, the algorithm
to classify facial expressions can be held in check by the spontaneous
facial expressions analysed.

Considering the above, the results provides an additional evidence that
individuals' subjective feeling can not be inferred from facial
expressions and in our case invalidate the hypothesis of hardwired
emotions. This result suggests that automatic facial expression
recognition tools should be focused on evaluating facial morphology
features such as action units rather than inferring potential emotional
or affective states.

>>>>

\hypersetup{
            pdftitle={Coherence between subjective feeling and facial expression: A comparison of human and automatic recognition},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Coherence between subjective feeling and facial expression: A comparison
of human and automatic recognition}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
While it has been taken for granted in the development of several
automatic facial expression recognition tools, the question of the link
between subjective feelings and facial expressions is still a subject of
debate. On one hand the behaviorist approach conceives emotions as
genetically hardwired and therefore being genuinely displayed through
facial expressions. On the other hand the constructivist approach
conceives emotions as socially constructed, the emotional meaning of a
facial expression being inferred by the observer. In order to evaluate
the link between the subjective feeling of emotions and their
recognition based on facial expression, 232 videos of encoders recruited
to carry out an emotion elicitation task were annotated by 1383 human
observers as well as by an automatic facial expression classifier.
Results show a low accuracy of human observers and of the automatic
classifier to infer the subjective feeling from the facial expressions
displayed by encoders . They also show a weak link between self-reported
emotional states and facial emotional displays. Based on these results,
the hypothesis of genetically hardwired emotion genuinely display is
difficult to support whereas the idea of emotion and facial expression
socially constructed appears to be more likely. Then automatic emotion
recognition tools based on facial expressions should be questioned.
\end{abstract}

@@

With the development of commercial automatic facial expression
recognition tools (see Dupré et al. 2018 for a non-exhaustive list of
available tools), industries and governments are gradually implementing
this technology in order to track humans' emotions in various scenarios
(e.g., marketing, healthcare, automotive to name a few). This
technoloogy rests on the premise that facial expressions provide a
direct access to individuals' subjective feeling. Even if this premise
is central to the modern mainstream approach of human emotion, recent
research in affective science are challenging it. Once the two competing
approaches briefly described as well as the predictions they
respectively entail, an experiment testing these hypotheses will be
presented and its results analyzed in order to provide empirical
evidence to contribute to answer the question.

\hypertarget{the-link-between-emotion-and-facial-expression-through-the-behaviorist-approach}{%
\subsection{The link between emotion and facial expression through the
behaviorist
approach}\label{the-link-between-emotion-and-facial-expression-through-the-behaviorist-approach}}

Based on the behaviorist approach initiated by Darwin in \emph{The
Expression of the Emotions in Man and Animals} (Darwin 1872), facial
expressions are conceived as a genuine displays of individuals' inner
emotional state. This hypothesis is used as a basis for the Basic
Emotion Theory (BET) which states that a set of six emotions are
universally displayed and are genetically hardwired not only in humans
(Ekman 1992) but also in different animal species (Waal 2019). According
to this view, ``when emotions are aroused by perception of a social
event, a set of central commands produce patterned emotion-specific
changes in multiple systems, including {[}\ldots{}{]} facial
expressions.'' (Ekman 2007, p49). To cope with critics, several
amendments have been made to the BET, increasing the number of basic
emotions from six to seven (Ekman and Heider 1988) as well as adding the
concept of ``display rules'' to explain cultural differences in the
management of facial expressions (Ekman et al. 1987).

Even if this theory obtained a popular support, it fails to explain how
individuals can feel emotions without expressing them and how
individuals can express emotions without feeling them, in instances in
which display rules cannot be called upon (Kraut and Johnston 1979;
Durán, Reisenzein, and Fernández-Dols 2017). These evidences have led to
an alternative conception, notably the social constructivist approach.

\hypertarget{the-link-between-emotion-and-facial-expression-through-the-social-constructivist-approach}{%
\subsection{The link between emotion and facial expression through the
social constructivist
approach}\label{the-link-between-emotion-and-facial-expression-through-the-social-constructivist-approach}}

Detractors of the Basic Emotion Theory consider emotion not as
genetically hardwired but as a learnt association between a given
situation and an appropriate response (Averill 1980; Barrett 2017a). For
the tenants of the constructivist approach, emotions are ``concepts''
based on past experiences and which are ``a collection of embodied,
whole brain representations that predict what is about to happen in the
sensory environment, what the best action is to deal with impending
events, and their consequences for allostasis'' (Barrett 2017b, p12).
Following this assumption, faces are best conceived as tools displaying
signals in social interactions (Crivelli and Fridlund 2018). These
signals can convey individuals' motivations and readiness (Frijda and
Tcherkassof 1997) or social messages (Fridlund and Rosenberg 1995).
Therefore, facial expressions are thought as behaviors which meaning is
inferred by the observer. Findings support this observer dependence
(Lindquist and Gendron 2013; Niedenthal et al. 2017). They show that to
make meaning of another person's facial behavior, the perceiver relies
in particular on her/his knowledge about emotion categories.

\hypertarget{the-recognition-of-emotional-facial-expressions}{%
\subsection{The recognition of emotional facial
expressions}\label{the-recognition-of-emotional-facial-expressions}}

Regarding emotional facial expression (EFE) recognition, the behaviorist
approach assertions and the constructivist approach claims lead to two
divergent predictions. The foremost postulates that, when triggered,
each basic emotion is expressed by a prototypical face (non basic
emotions being blends of the basic ones). Basic emotions are easily
recognized by all human observers and emotional states are accessible by
facial measurement. As the recognition of facial expressions is based on
the identification of specific patterns of facial movements, it implies
that EFE recognition of both human observers and automatic classifiers
should be as accurate. The constructivist approach, in contrast, affirms
that facial expressions do not provide a direct access to individuals'
subjective feeling. Even if the face possibly moves during an emotional
episode, facial muscle movements are not linked in a one-to-one manner
to a specific discrete emotional experience (Doyle and Lindquist 2017,
p418). Instead, emotions are mentally constructed by the perceiver and
mental categories of emotions are needed to accurately categorize facial
movements. Therefore, since there is no emotional prototypical face, one
should expect human observers' superior ability to accurately recognize
EFE as compared to automatic EFE recognition tools.

The aim of the current paper is to investigate the link between the
subjective feeling of emotions and its recognition from facial
expressions in order to give credit either to the behaviorist plea or to
the constructivist's one. Contrary to most studies using posed and
static EFE, the present study focuses on natural EFE. Spontaneous and
dynamic facial reactions to emotional elicitations are under
consideration to ensure the generalizability of the results to emotional
behaviors in ordinary life.

@@

To evaluate the link between subjective feeling of emotions and their
recognition from facial expressions, encoders were first recruited to
perform an emotion elicitation task while their facial expression was
video recorded. Then, the videos of the encoders' faces were shown to
human observers and were also analysed by an automatic classifier in
order to identify which emotion was displayed.

@@

For the emotion elicitation experiment, 358 French participants (182
females, 176 males, \emph{M}age = 47.9, \emph{SD}age = 9.2) were
recruited to perform one out of 11 emotion elicitation tasks designed to
trigger a positive, a specific negative or a neutral emotional state.
Encoders' face were recorded using an hidden camera resulting 358 front
facing 768x576 videos varying from 1s to 1479s (Figure
\ref{fig:dynemo_img}). These recordings form the DynEmo database (see
Tcherkassof et al. 2013 for a full description of tasks and procedure).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/dynemo_img-1.pdf}
\caption{\label{fig:dynemo_img}Example of a front facing recording
synced with the full view of the participant and the elicitation task.
This picture is taken from a pilot with projects collaborators and all
gave a consent for the publication of their photos and videos.}
\end{figure}

After the emotion elicitation task the encoders rated their subjective
feeling on Likert scales from 0 (``not at all'') to \nolinebreak 5
\nolinebreak (``strongly'') related to six ``basic'' emotion labels
(i.e., \emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) as well as six ``non-basic'' emotion
labels (i.e., \emph{pride}, \emph{curiosity}, \emph{boredom},
\emph{shame}, \emph{humiliation}, and \emph{disappointment}).

Finally, a debriefing session was performed to ensure that encoders were
not durably affected by the emotion elicitation task. The debriefing was
also used to check that encoders did not guess the real purpose of the
experiment (e.g.~being filmed while they were performing an emotional
elicitation task) to guarantee facial expressions' genuineness. All
encoders gave their agreement on their data and video to be processed
for research purpose only.

@@

For the human facial expression recognition method, 1383
\nolinebreak student participants were recruited to annotate 232 out of
the 358 videos, therefore only the 232 annotated videos will be analysed
in this paper. Because videos have different durations, participants had
to annotate a series of video corresponding to 30min long in total. Each
video was annotated 29 times on average (\emph{SD} = 12).

The annotation of facial expressions was performed on-site using
\emph{Oudjat}, a software for designing video annotation experiments
(Dupré et al. 2015). For each video, the annotation procedure followed
two steps. First, the participants had to identify the emotional
sequences by pressing the space bar of their keyboard to indicate the
beginning and the end of the emotional sequences while watching the
video. Second, the participants watched each emotional sequence
previously identified and labeled the sequence using one of the 12
\nolinebreak emotions proposed including six ``basic'' emotion labels
(i.e., \emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) and six ``non-basic'' emotion labels
(i.e., \emph{pride}, \emph{curiosity}, \emph{boredom}, \emph{shame},
\emph{humiliation}, and \emph{disappointment}). They also had the
possibility to indicate that the sequence was expressing none of the
proposed emotion labels.

This annotation procedure results in a uni-dimensional time-series for
each video per human observer identifying for each second of the video
which emotion was recognized. Then, time-series corresponding to the
same video were aggregated to calculate the proportion of human
observers \(x_{video_{i}.label_{j}.t_{k}}\) for each second of the video
per emotional label (EQ\ref{eq:1}).

\begin{equation}
\label{eq:1}
x_{video_{i}.label_{j}.t_{k}} = \frac{n_{video_{i}.label_{j}.t_{k}}}{n_{video_{k}}}
\end{equation}

where \emph{i} is one of the 232 videos, \emph{j} is one of the six
``basic'' emotion labels, \emph{k} for each second of the video.

@@

The 232 annotated video were processed with Affdex (SDK v3.4.1). Affdex
is an automatic facial expression recognition classifier developed and
distributed by Affectiva is a spin-off company resulting from the
research activities of MIT media lab created in 2009 (McDuff et al.
2016). Affdex's algorithm uses Histogram of Oriented Gradient (HOG)
features and Support Vector Machine (SVM) classifiers in order to
recognize facial expressions. For each video frame, Affdex identifies
the probability \(p_{video_{i}.label_{j}.t_{k}}\) from 0 to 100
(rescaled to 0 to 1 for the analysis) of the face as expressing one of
the six ``basic'' emotion labels (i.e., \emph{anger}, \emph{disgust},
\emph{fear}, \emph{happiness}, \emph{surprise} and \emph{sadness}) as
well as additional psychological states such as \emph{valence},
\emph{engagement} or \emph{contempt}, and facial features such as
\emph{cheek raise}, \emph{eye widen} or \emph{jaw drop}.

For both human and automatic recognition, to determine which of the six
``basic'' emotions can be used to identify each video, the recognition
probability for each label by frame was converted into odd ratio by
label (Dente et al. 2017). The highest sum of each odd ratio time-series
defines the label recognized by the automatic classifier (EQ\ref{eq:2}
for human recognition and EQ\ref{eq:3} for automatic recognition).
\begin{equation}
\label{eq:2}
video_{i}.label = \max\left(\frac{\sum_{k=1}^{n}x_{video_{i}.label_{j}.t_{k}}}{\sum_{k=1}^{n}x_{video_{i}.t_{k}}}\right)
\end{equation}

\begin{equation}
\label{eq:3}
video_{i}.label = \max\left(\frac{\sum_{k=1}^{f}p_{video_{i}.label_{j}.t_{k}}}{\sum_{k=1}^{f}p_{video_{i}.t_{k}}}\right)
\end{equation}

where \emph{i} is one of the 232 videos, \emph{j} is one of the six
``basic'' emotion labels, \emph{k} for each second of the \emph{n}
second video or for each frame of the \emph{f} frame video.

@@

Since encoders' self-reports, human annotations and the automatic
recognition include data on ``non-basic'' emotion labels and features,
the analysis is performed using only the six ``basic'' emotion labels in
order to compare them. The maximum score for self-reports, human
annotations and automatic recognition is used to label the video. In
case of more than one label obtaining the maximum value, the video is
labeled as undetermined.

@@

Encoders' subjective feeling is compared with human observers
recognition in a confusion matrix (Figure
\ref{fig:confusionMatrix_sr_hr}). Each emotion label used to describe
encoders' self-reported subjective feeling (i.e., the label rated with
the highest value) is compared with the emotion labels which were rated
with the highest score by human observers.

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_hr-1.pdf}
\caption{\label{fig:confusionMatrix_sr_hr}Confusion matrix of between
the emotion self-reported as being characteristic of the elicitation
with the emotion recognized by the human observers.}
\end{figure}

Results of the confusion matrix show a moderate agreement between the
emotion felt by the encoder during the elicitation and the emotion
recognized by the human observers (Accuracy \nolinebreak =
\nolinebreak 0.43, 95\%CI{[}0.35,0.52{]}; Kappa = 0.19) except for
\emph{disgust} (100\% of the videos self-reported). Sensitivity,
specificity, precision and F1 score for each emotion reveals that
\emph{happiness} has the highest coherence ratio whereas \emph{sadness}
has the lowest coherence ratio between true positives and false
positives (Table \ref{table:confusionTable_sr_hr}). Interestingly human
observers seem to recognize \emph{surprise} expressed in videos where
\emph{anger}, \emph{fear} \emph{happiness} and \emph{sadness} was the
highest self-reported emotion (respectively 28.6\%, 33.3\%, 34.7\% and
33.3\% of the videos self-reported), and in a lower instance
\emph{happiness} was recognized in videos where \emph{fear} and
\emph{surprise} was the highest self-reported emotion (respectively
50.0\% and 25.6\% of the videos self-reported).

@@

\caption{\label{tab:confusionTable_sr_hr}\label{table:confusionTable_sr_hr}Agreement accuracy metrics for each emotion.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
Emotion & Sensitivity & Specificity & Precision & F1\\
\midrule
anger & 0.43 & 0.93 & 0.23 & 0.3\\
disgust & 1.00 & 0.94 & 0.33 & 0.5\\
fear & 0.00 & 0.97 & 0.00 & \textit{na.}\\
happiness & 0.49 & 0.78 & 0.69 & 0.57\\
sadness & 0.08 & 0.92 & 0.08 & 0.08\\
surprise & 0.44 & 0.67 & 0.37 & 0.4\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} Note. \textit{na.} values are produced when not enough}\\
\multicolumn{5}{l}{data are available to compute accuracy indicators.}\\
\end{tabu}
\end{table}

However, self-reports show a significant proportion of undetermined
emotional states (35.2\% of the 358 videos) which reveals the potential
limit of using 6-points likert scales to measure emotional self-reports.
Indeed, encoders can easily score to the maximum for more than one
emotion.

@@

Similarly to the previous analysis, a confusion matrix was used to
compare encoders' subjective feeling with the emotion label recognized
by the automatic classifier (Figure \ref{fig:confusionMatrix_sr_ar}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_ar-1.pdf}
\caption{\label{fig:confusionMatrix_sr_ar}Confusion matrix of between
the emotion self-reported as being characteristic of the elicitation
with the emotion recognized by the automatic classifier.}
\end{figure}

Results obtained for the comparison between emotions self-reported and
recognized by the automatic classifier are somewhat similar to the ones
with human observers (Table \ref{table:confusionTable_sr_ar}). Overall,
a low agreement between emotion self-reported and emotion recognized by
the automatic classifier (Accuracy \nolinebreak = \nolinebreak 0.3,
95\%CI{[}0.22,0.38{]}; Kappa = 0.07) except for \emph{happiness} (44.4\%
of the video self-reported) is watched. Surprisingly the automatic
classifier incorrectly recognized as \emph{disgust} an important
proportion of videos in which \emph{anger}, \emph{happiness} and
\emph{surprise} was the highest self-reported emotion (respectively
42.9\%, 25.0\% and 32.6\% of the videos self-reported). In parallel, the
automatic classifier recognized as \emph{happiness} videos in which
\emph{fear} and \emph{surprise} was the highest self-reported emotion
(respectively 66.7\% and 32.6\% of the videos self-reported).

@@

\caption{\label{tab:confusionTable_sr_ar}\label{table:confusionTable_sr_ar}Agreement accuracy metrics for each emotion.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
Emotion & Sensitivity & Specificity & Precision & F1\\
\midrule
anger & 0.00 & 0.90 & 0.00 & \textit{na.}\\
disgust & 0.25 & 0.72 & 0.02 & 0.04\\
fear & 0.00 & 0.94 & 0.00 & \textit{na.}\\
happiness & 0.44 & 0.73 & 0.62 & 0.52\\
sadness & 0.15 & 0.95 & 0.25 & 0.19\\
surprise & 0.19 & 0.86 & 0.36 & 0.25\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} Note. \textit{na.} values are produced when not enough}\\
\multicolumn{5}{l}{data are available to compute accuracy indicators.}\\
\end{tabu}
\end{table}

A comparable explanation involing the amount of undertermined video
based on self-reports can be provided as the level of undetermined
emotions are very high for the self reports.

\hypertarget{comparison-between-human-and-automatic-recognition}{%
\subsection{Comparison between human and automatic
recognition}\label{comparison-between-human-and-automatic-recognition}}

As previously mentioned, human observers appear to be more accurate than
the automatic classifier to recognize individuals' subjective feeling
(human observer Accuracy = 0.43; automatic classifier Accuracy = 0.3).
However, both make mistakes. A third confusion matrix is used to compare
similarities (diagonal) and differences between human observers and
automatic classifier in classifing the six emotion labels (Figure
\nolinebreak \ref{fig:confusionMatrix_hr_ar}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_hr_ar-1.pdf}
\caption{\label{fig:confusionMatrix_hr_ar}Comparison between the human
observers and automatic classifier recognition for the classfication of
emotion labels.}
\end{figure}

The overall agreement between human observers and the automatic
classifier is in fact very low (Kappa = 0.18). Except for
\emph{happiness} and \emph{disgust} (respectively 64.9\% and 50.0\% of
common labelling), there is no clear common pattern. Moreover, the
automatic classifier has a tendency to label as \emph{disgust} videos
labeled as \emph{sadness} by human observers and as \emph{happiness}
videos labaled as \emph{fear} by human observers

\hypertarget{discussion-and-conclusion}{%
\section{Discussion and Conclusion}\label{discussion-and-conclusion}}

Despite being one on the most investigated question in affective
science, the coherence between emotion felt and facially displayed and
facial expression recognized is a hot topic. To date no clear evidence
has been found to definitely answer it. Yet, with the growing interest
of industries and government to monitor individual's psychological
states, this issue is under high pressure. The present research aims to
provide some empirical data to the question. The faces encoders
displayed when confronted with an emotional eliciting task were
submitted both to human and to automatic recognition. The criterion for
recognition accuracy was the subjective feeling self-reported by the
encoder once the elicitation task carried out. Results first reveal a
low coherence between emotion felt and facial expression displayed.
Secondly, results show a low accuracy for both humans and automatic
classifier in identifying the inner emotional states of these encoders
based on their facial expressions. Thirdly, human observers prove to be
better at recognizing the emotion facially expressed than the automatic
recognition tool is.
